Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-23 articles/cycle/response-monegato.tex

    \msquote{%
        Finally, modern precipitation data from WorldClim also bear
        uncertainties and exhibit local disagreement with other regional data
        \citep{Isotta.etal.2013}.}

    Besides, the following sentence was added to the discussion of glacier
    extent in the south-western Alps,

    \msquote{%
    However, geochronological data from the south-western Alps are sparse, and
    the LGM extent compilation \citep{Ehlers.Gibbard.2004} is inconsistent with
    more recent regional reconstructions \citep{Federici.etal.2017}.}

    \referee{\textbf{2 --}
        Concerning the ice-transfluence. I agree that it has to be much more
        considered. I wonder if the Adige had a great contribution from the
        Austrian Tauern (Toblach area) this would have increased the ice in
        Adige and deplete the Drava, or in Winschgau valley where the catchment
        upstream Mustar saddle is more related to Adige than the Inn. The same
        could have happened for the Piave catchment, which is very
        underestimated in the model. Again, ice flowing to the south did not
        flow to the east. But I think that this may be not enough for
        justifying the overestimation of the Eastern Alps. Concerning the
        Straninger saddle I am a bit skeptic because it is not the lowermost
        saddle (Plockenpass and Nassfeldpass are lower in elevation). The
        central Adda glacier (and not the Ticino as at page 14 line 26) is much
        underestimated and here the contribution from transfluence in St.
        Moritz area could have been remarkable. I attach a figure separately
        with transfluences (yellow stars and red arrows).}

    The amount of ice flowing through transfluences is certainly dependent on
    model resolution and basal sliding parameters. However, we agree that this
    will have little effect on the ice volume overestimation in the eastern
    Alps in comparison to potential shortcomings in the climate forcing.

    The tranfluence at the Straniger Saddle may be a byproduct of bedrock
    topography aggregation to the 1\,km grid.  As a follow-up study, we have
    begun to use higher model resolution to study the location of transfluences
    in more detail \citep{Seguinot.etal.2018a}. In these results, the
    transfluence at the Straniger Saddle looses importance relative to other
    transfluences in the region including, in fact, Plöckenpass and
    Nassfeldpass. Because the choice of transfluences plotted on Fig.~4 is
    somewhat arbitrary, we have decided to update it.

### 2018-07-23 articles/cycle/response-monegato.tex

    The amount of ice flowing through transfluences is certainly dependent on
    model resolution and basal sliding parameters. However, we agree that this
    will have little effect on the ice volume overestimation in the eastern
    Alps in comparison to potential shortcomings in the climate forcing.

    The tranfluence at the Straniger Saddle may be a byproduct of bedrock
    topography aggregation to the 1\,km grid.  As a follow-up study, we have
    begun to use higher model resolution to study the location of transfluences
    in more detail \citep{Seguinot.etal.2018a}. In these results, the
    transfluence at the Straniger Saddle looses importance relative to other
    transfluences in the region including, in fact, Plöckenpass and
    Nassfeldpass. Because the choice of transfluences plotted on Fig.~4 is
    somewhat arbitrary, we have decided to update it.

    \referee{\textbf{3 --}
        The basal sliding of glaciers and its overall velocity is another
        interesting factor to discuss. For Adige for example we know that at
        about 28 cal BP the Adige glacier was damming a tributary valley north
        of Trento around the same age (Avanzini et al 2009), while the same
        glacier arrived at Garda at 24.6 cal BP (Monegato et al., 2017).  This
        means that the glacier front advanced of about 100 km in around 3.5 ka
        so 280 m/y. The same could have been for the Adda and Oglio glaciers,
        even if a such robust chronology is lacking. Is it possible that
        overestimated velocities produced large eastern glaciers?}

    Thank you. In fact basal sliding is one of the major uncertainties in our
    results which is now better acknowledged for in our manuscript following
    other reviewer's comments. However, basal sliding velocity are somewhat
    decoupled from glacier front advance velocities and glacier front maximum
    extent. Although this is not shown in our manuscript, increased basal
    velocities tend to result in thinner ice tongues. Due to the
    mass-balance-elevation feedback, this causes more surface melt and, in
    turn, lesser extent. However this effect is small in comparison to changes
    due to climate forcing.

### 2018-07-23 articles/cycle/response-monegato.tex

    \referee{\textbf{3 --}
        The basal sliding of glaciers and its overall velocity is another
        interesting factor to discuss. For Adige for example we know that at
        about 28 cal BP the Adige glacier was damming a tributary valley north
        of Trento around the same age (Avanzini et al 2009), while the same
        glacier arrived at Garda at 24.6 cal BP (Monegato et al., 2017).  This
        means that the glacier front advanced of about 100 km in around 3.5 ka
        so 280 m/y. The same could have been for the Adda and Oglio glaciers,
        even if a such robust chronology is lacking. Is it possible that
        overestimated velocities produced large eastern glaciers?}

    Thank you. In fact basal sliding is one of the major uncertainties in our
    results which is now better acknowledged for in our manuscript following
    other reviewer's comments. However, basal sliding velocity are somewhat
    decoupled from glacier front advance velocities and glacier front maximum
    extent. Although this is not shown in our manuscript, increased basal
    velocities tend to result in thinner ice tongues. Due to the
    mass-balance-elevation feedback, this causes more surface melt and, in
    turn, lesser extent. However this effect is small in comparison to changes
    due to climate forcing.

    \referee{\textbf{4 --}
        About the self-sustained ice domes, their importance is for me unclear
        and not well explained in the text. Why they are only two? Why the
        Adamello or the Tauern massifs, as an example, were not considered as
        self-sustained ice domes?}

    By self-sustained ice domes, we mean ice domes that are not sitting on top
    of basal topographic highs, but instead shift away from the modern
    topographic divides and ``sustain'' themselves by reaching higher
    elevation than the local mountains. We indicate these as a sign that ice
    begins to behave as in an ice sheet, and flow in directions independent
    from the local basal topography.

    However, we agree that this is somewhat arbitrary, and again, as we found
    out, dependent on model resolution \citep{Seguinot.etal.2018a}. Therefore,
    we decided to remove the ``ice domes'' from Fig.~4a and the main text, and
    save this discussion for future studies.

### 2018-07-23 articles/cycle/response-monegato.tex

    By self-sustained ice domes, we mean ice domes that are not sitting on top
    of basal topographic highs, but instead shift away from the modern
    topographic divides and ``sustain'' themselves by reaching higher
    elevation than the local mountains. We indicate these as a sign that ice
    begins to behave as in an ice sheet, and flow in directions independent
    from the local basal topography.

    However, we agree that this is somewhat arbitrary, and again, as we found
    out, dependent on model resolution \citep{Seguinot.etal.2018a}. Therefore,
    we decided to remove the ``ice domes'' from Fig.~4a and the main text, and
    save this discussion for future studies.

    \referee{\textbf{5 --}
        Global circulation models (e.g., Löfverström et al., 2014, Beghin et
        al., 2015) suggest that the Polar front was at different latitudes
        during each cold phases of the Late Pleistocene. This could have effect
        of different impact on the Alps. For example during MIS4 if the
        westerlies were dominant this could have driven more effective
        precipitations in the western and northern Alps in respect to the
        southern and the eastern Alps. If this is true, and can be applied to
        the major cold phases. How was the behavior of glaciers during MIS 5
        and 3?}

    We agree. In fact, our simulations are limited by the constant temperature
    and precipitation patterns applied throughout the glacial cycle. Although
    we can't answer this question without new climate and glacier simulations,
    we have raised this important point in the very last paragraph:

    \msquote{%
        Shifts in the North Atlantic storm track and polar front may have
        caused varied patterns of glaciations through different cold phases.}

    With this, we would like to thank you again very much for the time and
    effort you put into our manuscript.

### 2018-07-23 articles/cycle/response-monegato.tex

    Thank you for this detailed analysis. We want to point out that WorldClim
    is not a model, but is also based on observations. However we agree, that
    the regional dataset by \citet{Isotta.etal.2013}, tailored to the dense
    station network and steep topography of the Alps, is certainly a more
    robust precitation forcing that needs to be considered in future
    studies. A reference was added:

@@

    \msquote{%
        Finally, modern precipitation data from WorldClim also bear
        uncertainties and exhibit local disagreement with other regional data
        \citep{Isotta.etal.2013}.}

    Besides, the following sentence was added to the discussion of glacier
    extent regarding the south-western Alps,

@@

    \referee{\textbf{5 --}
        Global circulation models (e.g., Löfverström et al., 2014, Beghin et
        al., 2015) suggest that the Polar front was at different latitudes
        during each cold phases of the Late Pleistocene. This could have effect
        of different impact on the Alps. For example during MIS4 if the
        westerlies were dominant this could have driven more effective
        precipitations in the western and northern Alps in respect to the
        southern and the eastern Alps. If this is true, and can be applied to
        the major cold phases. How was the behavior of glaciers during MIS 5
        and 3?}

    We agree. In fact, our simulations are limited by the constant temperature
    and precipitation patterns applied throughout the glacial cycle. Although
    we can't answer this question without new climate and glacier simulations,
    we have raised this important point in the very last paragraph:

    \msquote{%
        Shifts in the North Atlantic storm track and polar front may have
        caused varied patterns of glaciations through different cold phases.}

    With this, we would like to thank you again very much for the time and
    effort you have put into our manuscript.

### 2018-07-24 articles/cycle/alps-cycle.tex
Address Giovanni's comment on polar front.
Replace Straniger Alm transfluence in the text.

    \begin{figure*}[t]
      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
      \caption{%
        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
        (200\,m contours), and ice surface velocity (blue) in the Alps
        24.57\,ka before present, corresponding to the maximum modelled ice
        cover. The modelled LGM ice extent (dashed orange line) and
        geomorphological reconstruction \citep[solid red
        line,][]{Ehlers.etal.2011}, modelled self-sustained ice domes
        (triangles), and some major transfluences (crosses) are shown.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model \citep[black
        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
        last glacial cycle (120--0\,ka), expressed in meters of sea level
        equivalent (m~s.l.e., blue curve). Shaded gray areas indicate the
        timing for Marine Oxygen Isotope Stage (MIS)~2 and MIS~4 according to
        a~global compilation of benthic \chem{\delta^{18}O} records
        \citep{Lisiecki.Raymo.2005}.
        \todo{Thicken the 200\,m contours on all figures.}}
      \label{fig:lgmvel}
    \end{figure*}

@@

    In the Eastern Alps, major transfluences are modelled to have occurred for
    instance across Fern Pass, the Seefeld Saddle, the Gailberg Saddle, the
    Kreuxberg Saddle, the Straniger Alp, and Pyhrn Pass
    (Fig.~\ref{fig:lgmvel}a). Transfluences across Fern Pass and the Seefeld
    Saddle are known from the geomorphology
    \citep[Fig.~2.4]{Penck.Bruckner.1909, Husen.2011}. It is also known that
    ice flowed across the Gailberg and Kreuzberg saddles \citep{Husen.1985},
    and Pyhrn Pass \citep[Fig.~2.5]{Husen.2011}.
    On the other hand, no transfluence has previously been documented at the
    Straniger Alp. The Tagliamento catchment is usually assumed to not have
    received tranfluences from the Drava catchment \citep{Monegato.etal.2007},
    but this would not be incompatible with reconstructed ice surface
    elevations in this area \citep{Husen.1987}.

@@

    More importantly, these results are limited by the simplicity of the
    surface mass balance parameters and climate forcing used.
    In particular, additional palaeo-climate variability over
    the European Alps may have caused more glaciations than modelled here.
    Using more specifically targeted sensitivity runs, and a more realistic
    climate forcing based on regional circulation model output or including
    the effect of long-term term changes in incoming solar radiation, future
    modelling studies will certainly be able to quantify uncertainties
    associated with some of the above limitations. Nevertheless, we hope that
    these conclusions will also serve as a basis for future studies of glacial
    geology in the Alps, and call for a more systematic aggregation and
    homogenization of glacial geological data to form a basis for model
    validation across the entire Alpine range.

>>>>

    \begin{figure*}[t]
      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
      \caption{%
        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
        (200\,m contours), and ice surface velocity (blue) in the Alps
        24.57\,ka before present, corresponding to the maximum modelled ice
        cover. The modelled LGM ice extent (dashed orange line) and
        geomorphological reconstruction \citep[solid red
        line,][]{Ehlers.etal.2011}, modelled self-sustained ice domes
        (triangles), and some major transfluences (crosses) are shown.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model \citep[black
        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
        last glacial cycle (120--0\,ka), expressed in meters of sea level
        equivalent (m~s.l.e., blue curve). Shaded gray areas indicate the
        timing for Marine Oxygen Isotope Stage (MIS)~2 and MIS~4 according to
        a~global compilation of benthic \chem{\delta^{18}O} records
        \citep{Lisiecki.Raymo.2005}.
        \todo{Thicken the 200\,m contours on all figures. Remove ice domes and
              replace Straniger.}}
      \label{fig:lgmvel}
    \end{figure*}

@@

    In the Eastern Alps, major transfluences are modelled to have occurred for
    instance across Fern Pass, the Seefeld Saddle, the Gailberg Saddle, the
    Kreuzberg Saddle, Kronhof Pass, Studenbodenalm, and Pyhrn Pass
    (Fig.~\ref{fig:lgmvel}a). Transfluences across Fern Pass and the Seefeld
    Saddle are known from the geomorphology
    \citep[Fig.~2.4]{Penck.Bruckner.1909, Husen.2011}. It is also known that
    ice flowed across the Gailberg and Kreuzberg saddles \citep{Husen.1985},
    and Pyhrn Pass \citep[Fig.~2.5]{Husen.2011}.
    On the other hand, no transfluence has previously been documented over the
    Kronhof Pass, Studenbodenalm or elsewhere over the Karnic Alps.
    The Tagliamento catchment is usually assumed to not have
    received tranfluences from the Drava catchment \citep{Monegato.etal.2007},
    but this would not be incompatible with reconstructed ice surface
    elevations in this area \citep{Husen.1987}.

@@

    More importantly, these results are limited by the simplicity of the
    surface mass balance parameters and climate forcing used.
    In particular, additional palaeo-climate variability over
    the European Alps may have caused more glaciations than modelled here.
    Shifts in the North Atlantic storm track and polar front may have caused
    varied patterns of glaciations through different cold phases.
    Using more specifically targeted sensitivity runs, and a more realistic
    climate forcing based on regional circulation model output or including
    the effect of long-term term changes in incoming solar radiation, future
    modelling studies will certainly be able to quantify uncertainties
    associated with some of the above limitations. Nevertheless, we hope that
    these conclusions will also serve as a basis for future studies of glacial
    geology in the Alps, and call for a more systematic aggregation and
    homogenization of glacial geological data to form a basis for model
    validation across the entire Alpine range.
